Toward a secure TCP/IP stack --- Technical report

### 2020-07-24 report/report.tex
Update report

@@

    CycloneTCP is a library developped by the company Oryx Embedded for embedded platforms.
    A large number of platforms are supported, and the library can be used
    with a dozen of OS. The code of the library is written in C and implements
    many protocols from low-level network layers like IPv4/IPv6, to transport layers
    and application layers like DHCP or HTTP. In particular an implementation of
    the TCP protocol is provided.

    SPARK 2014 is a programming language designed as a subset of Ada that helps in
    making high reliability sofware by providing powerful static analysis methods.
    SPARK is able to detect uninitialized variables with control flow analysis,
    and can ensure the absence of run-time errors,
    but also, based on SMT-solver, functional behavior can be specified for every
    function and mathematically proved.

@@

        %
        % Position States
        %
        \node[state, name=closedStart] {CLOSED};
        \node[state, name=listen, below of=closedStart] {LISTEN};
        \node[state, name=synSent, below of=listen, right of=listen, xshift=8em] {SYN\_SENT};
        \node[state, name=synRcvd, below of=listen, left of=listen, xshift=-8em] {SYN\_RCVD};
        \node[state, name=established, below of=listen, node distance=14em] {ESTABLISHED};
        \node[state, name=finWait1, below of=established, left of=established, node distance=7em, xshift=-9em] {FIN\_WAIT\_1};
        \node[state, name=finWait2, below of=finWait1] {FIN\_WAIT\_2};
        \node[state, name=closeWait, below of=established, right of=established, node distance=7em, xshift=9em] {CLOSE\_WAIT};
        \node[state, name=closing, below of=established, node distance=14em] {CLOSING};
        \node[state, name=lastAck, below of=closeWait] {LAST\_ACK};
        \node[state, name=timeWait, below of=closing] {TIME\_WAIT};

@@

        \draw (synSent.west) 
        edge[stateEdge] node[edgeLabel, yshift=1em]{SYN/SYN + ACK} 
        (synRcvd.east);
        \draw (synRcvd) 
        edge[stateEdge] node[edgeLabel, xshift=-2.5em]{\emph{Close}/FIN} 
        (finWait1);

        \draw ($(synRcvd.east) + (0,-1em)$) 
        edge[stateEdge, bend left=22.5] node[edgeLabel, xshift=-1em, yshift=-1em]{ACK} 
        ($(established.north) + (-1em,0)$);
        \draw ($(synSent.west) + (0,-1em)$) 
        edge[stateEdge, bend right=22.5] node[edgeLabel, xshift=3em, yshift=-1em]{SYN + ACK/ACK} 
        ($(established.north) + (1em,0)$);

@@

        \draw (finWait1.south) 
        edge[stateEdge] node[edgeLabel, xshift=-2em]{ACK} 
        (finWait2.north);
        \draw ($(finWait1.east) + (0,-.5em)$) 
        edge[stateEdge, bend left=22.5] node[edgeLabel, yshift=1em]{ACK} 
        (closing.north);
        \draw (finWait1.south east) 
        edge[stateEdge] node[edgeLabel, xshift=1em, yshift=2em, text width=3em]{FIN + ACK/ACK} 
        (timeWait.north west);

@@

        \draw (closing) 
        edge[stateEdge] node[edgeLabel, xshift=-2em]{ACK} 
        (timeWait);

@@

    Different states can be taken by a socket during its lifetime depending on the state
    of the connection. The state also gives information about the state of the remote TCP.
    Each state has its own signification:
    \begin{description}
        \item[CLOSED] represents no connection state at all.
        \item[LISTEN] represents waiting for a connection request from any remote
        TCP and port.
        \item[SYN-SENT] represents waiting for a matching connection request
        after having sent a connection request.
        \item[SYN-RECEIVED] represents waiting for a confirming connection
        request acknowledgment after having both received and sent a
        connection request.
        \item[ESTABLISHED] represents an open connection, data received can be
        delivered to the user. The normal state for the data transfer phase
        of the connection.
        \item[FIN-WAIT-1] represents waiting for a connection termination request
        from the remote TCP, or an acknowledgment of the connection
        termination request previously sent.
        \item[FIN-WAIT-2] represents waiting for a connection termination request
        from the remote TCP.
        \item[CLOSE-WAIT] represents waiting for a connection termination request
        from the local user.
        \item[CLOSING] represents waiting for a connection termination request
        acknowledgment from the remote TCP.
        \item[LAST-ACK] represents waiting for an acknowledgment of the
        connection termination request previously sent to the remote TCP
        (which includes an acknowledgment of its connection termination
        request).
        \item[TIME-WAIT] represents waiting for enough time to pass to be sure
        the remote TCP received the acknowledgment of its connection
        termination request.
    \end{description}

    For more details on how two TCP can establish or close a connection, the reader can refer
    to the TCP norm where many scenarios are described and explained, and more
    details are given on the state and segments that can be sent.

    \subsubsection{Multithreading in the implementation}

    Different threads can interact with a socket. The TCP norm gives an example with
    three threads: one for the \emph{user calls}, one for the \emph{arriving segments}
    and one for the \emph{timers}. This design has also been adopted in CycloneTCP.

    \paragraph{User calls}
    User calls refer to functions that can be called by the user to control the connection,
    send, or received data: OPEN, CLOSE, ABORT, SEND, RECEIVED. Transition between states of
    the connection can happen during the call of these functions since they are intended to control
    the connection.

    \paragraph{Arriving segments}
    In this thread, the received segments are processed. The corresponding messages are sent back.
    Transitions between states can happen here when a message is received. For example,
    this is the case when a message containing a SYN is received when the socket is in
    the LISTEN state. It will automatically send a SYN + ACK in response, and change
    its state for SYN-RECEIVED.

    \paragraph{Timers}
    The timers control the timeouts like the retransmission timeout to retransmit a message, or
    the time-wait timeout to close the connection after an amount of time. Then, transitions can also
    happen in this thread.

    To sum up, a single socket can be seen as a shared structure over multiple threads, that can all
    manipulate it and change the value of its fields. The schema in figure \ref{fig:TCPConcurrency} illustrates that.

    Only one thread can perform an operation on the socket at the same time. The access to the fields of
    the socket are protected by mutex.

    Two threads can communicate synchronously or asynchronously together. The synchronous communications are based on
    the interface provided by the OS, in particular with events.

@@

            \draw[decorate, decoration={brace, amplitude=10pt}] (7.5, 0) -- (7.5, -1.8)
                node [black,midway,xshift=0.55cm, rotate=-90] {\textit{Threads}};

@@

    The TCP norm defines a lot of behavior that could be verified by automatic proof.
    Here is a non exhaustive list of what could be done:
    \begin{itemize}
        \item Verification of the transitions between the states.
        \item Integrity of the messages sent, \textit{i.e.} we could check that the sent
        message contains the correct flag and the correct sequence and acknowledgement numbers.
        \item Integrity of the messages received, \textit{i.e.} we could check that the
        messages received are correctly processed in regard of the flags they contain.
        \item Functional correctness of the user functions in regard to their specifications.
    \end{itemize}

    Some choices have had to be done in regard to the short time devolved to the internship, and
    not every aspect of the TCP norm has been proved yet. However, some aspects have been selected
    for the challenge or the importance they represent. A list of these aspects is presented
    here and the choices will be motivated in the next subsections. First of all, we have wanted
    to improve the interface of the functions. The other point was to improve the safety of the
    protocol itself.

@@

    One of the main problems pointed to by the main author of the CycloneTCP library is that users
    do not correctly use the API supplied. As a result, some function
    calls can be incorrect, either because the arguments are incorrect, or because the
    return code of the previous calls has not been checked, which lead to an
    incorrect code.

@@

    The multiples tasks and the changes that can happen at every moment in one
    task or another make the verification hard. Multiples interactions exist
    between the multiples threads: synchronous and asynchronous.
    All these interactions have to be considered if we want to write correct contracts
    in regard of what could be done in other threads.

    As SPARK does not have a native mode to deal precisely with interactions related to concurrency, we have to
    model these different interactions by hands by writing assertions that
    modify the state of the socket as another thread could do it. In particular
    the problem has been encountered when we tried to write contracts for the correctness
    of the user functions. More details and explanations will be given in the following
    section, where we will investigate the solution found to these technical problems.

@@

    This part mainly focusses on the high level user functions, the socket functions.
    The socket functions are the one called by the user to perform operations on
    the network. They are located in the files \texttt{socket\_interface.ad(b|s)}.

@@

    In order to verify the transitions we have read the norm to extract informations about
    the allowed transitions, and we have added contracts in the function
    \lstinline[language=Ada]{tcpChangeState}. Since all the transitions are made through
    this function, an incorrect transition will lead to a message by SPARK.

    As mentioned, since most transitions are done outside the user functions, a big part
    of this work was to consider other transitions in other threads and we present how we have
    done it in the following sections.

@@

    Almost all the transitions are done in the file \texttt{tcp\_fsm.c}. This file has not been
    translated in SPARK by a lack of time. Then we do not have strong guaranties on its functions
    behavior. However, for formal verification we need to know what is done in the functions of
    this file, because a wrong contract can make all the verification wrong.

@@

    All the work related to KLEE is present in the folder \texttt{klee/} and can be compiled and run
    thanks to the makefile to reproduce the results. Roughtly, what is done is creating a random
    incoming message, put the socket in the desired state and call the desired function to see what
    final states can be obtain.

    Finally, all the results found by KLEE have been reported as a post-condition of the function \lstinline[language=Ada]{Tcp_Process_One_Segment}.
    This function is essential for all the verification of concurrent parts and is reused everywhere
    the concurrency is involved to compute possible interferences. It follows that the safety of the code rely on the
    confidence we have in this function.

@@

    Let $\rightarrow$ be the transition function in the TCP automaton restricted to the transitions
    that can be performed by the reception of a message and its automatic response mechanism.
    Then we have $\rightarrow = \text{\lstinline[language=Ada]{Tcp_Process_One_Segment}}$ and
    we need to consider the reflexive transitive closure $\rightarrow^*$ of $\rightarrow$, with
    \[\rightarrow^* = \bigcup_{n\in\mathbb{N}} \rightarrow^n\]
    Now, by examining the TCP automaton in figure \ref{fig:TCPAutomaton}, we see that all state is
    only reachable by a state at a distance less than 3 (the maximum path is between SYN-SENT and
    CLOSE-WAIT if we pass by state SYN-RECEIVED and ESTABLISHED) without user action. Since
    we only consider the transitive closure in term of states, we can significantly reduce the number
    of iteration of $\rightarrow$ to compute $\rightarrow^*$ and finally we get:
    \[\rightarrow^* = \bigcup_{n=1}^3 \rightarrow^n\]
    It follows that algorithm \ref{algo:tcpProcessSegment} is good enough to compute the transitive closure of the function
    \lstinline[language=Ada]{Tcp_Process_One_Segment} by taking advantage of the fact that SPARK can unroll
    small loops to obtain more precise results.

@@

    In the code, the function \lstinline[language=C]{tcpUpdateEvents} is called at different locations
    and can raise the desired event. In particular this function is called each time the state of
    the socket is changed. This is sufficient for our purpose since we are only interested in the state
    changes.

@@

    \begin{algorithm}[t]
        \SetKwFunction{ProcessSegm}{\lstinline[language=Ada]{Tcp\_Wait\_For\_Events}}
        \SetKwProg{Fn}{function}{}{end}
        \Fn{\ProcessSegm{Socket, Event, Event\_Mask}}{
            $S_{last} := \text{\textit{Socket}}$\;
            $S := S_{last}$\;
            $E :=$ \lstinline[language=Ada]{Tcp_Process_One_Segment}($S_{last}$)\;
            \If{$E \wedge \text{Event\_Mask} \neq 0$}{
                \KwRet{$S$}\;
            }
            \For{$i=1$ \KwTo $3$}{
                $S_{last} :=$ \lstinline[language=Ada]{Tcp_Process_One_Segment}($S_{last}$) \;
                $S := S \cup S_{last}$\;
                $E :=$ \lstinline[language=Ada]{Tcp_Process_One_Segment}($S_{last}$)\;
                \If{$E \wedge \text{Event\_Mask} \neq 0$}{
                    \KwRet{$S$}\;
                }
            }
            \KwRet{$\emptyset$}\;
        }
        \caption{Function to compute the possible state after when waiting for a particular event.}
        \label{algo:waitForEvents}
    \end{algorithm}

@@

    \textcolor{red}{\paragraph{Weakness} If two threads want to lock the same mutex at the same time, which one win?
    This is probably something to discuss with Clément, because the function \texttt{tcpWaitForEvents} could have a
    different semantic than what we imagine at first.}

@@

    Other improvements that could be done in the future:
    \begin{itemize}
        \item Fully translate the file \texttt{tcp\_misc.c} in Ada to have more guarantees.
        \item Same for the file \texttt{tcp\_fsm.c} that has been investigate with klee. But
        it is not enough and we need more guarantees for this file, in order to prove the
        interactions between threads first, and also to prove the correctness of this part
        of code w.r.t. the TCP norm.
    \end{itemize}

>>>>

\let\state\textsf
\newcommand\ESTABLISHED{\state{ESTABLISHED}}
\newcommand\SYNRECEIVED{\state{SYN-RECEIVED}}
\newcommand\CLOSED{\state{CLOSED}}
\newcommand\CLOSEWAIT{\state{CLOSE-WAIT}}
\newcommand\LISTEN{\state{LISTEN}}
\newcommand\SYNSENT{\state{SYN-SENT}}
\newcommand\FINWAITONE{\state{FIN-WAIT-1}}
\newcommand\FINWAITTWO{\state{FIN-WAIT-2}}
\newcommand\CLOSING{\state{CLOSING}}
\newcommand\LASTACK{\state{LAST-ACK}}
\newcommand\TIMEWAIT{\state{TIME-WAIT}}

@@

    CycloneTCP is a library developed by the company Oryx Embedded for embedded platforms.
    A large number of platforms are supported, and the library can be used
    with a dozen of OS. The code of the library is written in C and implements
    many protocols from low-level network layers like IPv4/IPv6, to transport layers
    and application layers like DHCP or HTTP. In particular an implementation of
    the TCP protocol is provided.

    SPARK 2014 is a programming language designed as a subset of Ada that helps in
    making high reliability software by providing powerful static analysis methods.
    SPARK is able to detect uninitialized variables with control flow analysis,
    and can ensure the absence of run-time errors,
    but also, based on SMT-solver, functional behavior can be specified for every
    function and mathematically proved.

@@

        %
        % Position States
        %
        \node[state, name=closedStart] {CLOSED};
        \node[state, name=listen, below of=closedStart] {LISTEN};
        \node[state, name=synSent, below of=listen, right of=listen, xshift=8em] {SYN\_SENT};
        \node[state, name=synRcvd, below of=listen, left of=listen, xshift=-8em] {SYN\_RECEIVED};
        \node[state, name=established, below of=listen, node distance=14em] {ESTABLISHED};
        \node[state, name=finWait1, below of=established, left of=established, node distance=7em, xshift=-9em] {FIN\_WAIT\_1};
        \node[state, name=finWait2, below of=finWait1] {FIN\_WAIT\_2};
        \node[state, name=closeWait, below of=established, right of=established, node distance=7em, xshift=9em] {CLOSE\_WAIT};
        \node[state, name=closing, below of=established, node distance=14em] {CLOSING};
        \node[state, name=lastAck, below of=closeWait] {LAST\_ACK};
        \node[state, name=timeWait, below of=closing] {TIME\_WAIT};

@@

        \draw (synSent.west) 
        edge[stateEdge] node[edgeLabel, yshift=1em]{SYN/ACK} 
        (synRcvd.east);
        \draw (synRcvd) 
        edge[stateEdge] node[edgeLabel, xshift=-2.5em]{\emph{Close}/FIN} 
        (finWait1);

        \draw ($(synRcvd.east) + (0,-1em)$) 
        edge[stateEdge, bend left=22.5] node[edgeLabel, xshift=-1em, yshift=-1em]{ACK/-} 
        ($(established.north) + (-1em,0)$);
        \draw ($(synSent.west) + (0,-1em)$) 
        edge[stateEdge, bend right=22.5] node[edgeLabel, xshift=3em, yshift=-1em]{SYN + ACK/ACK} 
        ($(established.north) + (1em,0)$);

@@

        \draw (finWait1.south) 
        edge[stateEdge] node[edgeLabel, xshift=-2em]{ACK/-} 
        (finWait2.north);
        \draw ($(finWait1.east) + (0,-.5em)$) 
        edge[stateEdge, bend left=22.5] node[edgeLabel, yshift=1em]{FIN/ACK} 
        (closing.north);
        \draw (finWait1.south east) 
        edge[stateEdge] node[edgeLabel, xshift=1em, yshift=2em, text width=3em]{FIN + ACK/ACK} 
        (timeWait.north west);

@@

        \draw (closing) 
        edge[stateEdge] node[edgeLabel, xshift=-2em]{ACK/-} 
        (timeWait);

@@

    Different states can be taken by a socket during its lifetime depending on the state
    of the connection. The state also gives information about the state of the remote TCP.
    Each state has its own signification:
    \begin{description}
        \item[\CLOSED] represents no connection state at all.
        \item[\LISTEN] represents waiting for a connection request from any remote
        TCP and port.
        \item[\SYNSENT] represents waiting for a matching connection request
        after having sent a connection request.
        \item[\SYNRECEIVED] represents waiting for a confirming connection
        request acknowledgment after having both received and sent a
        connection request.
        \item[\ESTABLISHED] represents an open connection, data received can be
        delivered to the user. The normal state for the data transfer phase
        of the connection.
        \item[\FINWAITONE{}] represents waiting for a connection termination request
        from the remote TCP, or an acknowledgment of the connection
        termination request previously sent.
        \item[\FINWAITTWO{}] represents waiting for a connection termination request
        from the remote TCP.
        \item[\CLOSEWAIT] represents waiting for a connection termination request
        from the local user.
        \item[\CLOSING{}] represents waiting for a connection termination request
        acknowledgment from the remote TCP.
        \item[\LASTACK{}] represents waiting for an acknowledgment of the
        connection termination request previously sent to the remote TCP
        (which includes an acknowledgment of its connection termination
        request).
        \item[\TIMEWAIT{}] represents waiting for enough time to pass to be sure
        the remote TCP received the acknowledgment of its connection
        termination request.
    \end{description}

    \paragraph{Example of a possible way to establish a connection}

    We present a scenario on how a connection can be established between two TCPs,
    when one TCP (TCP A) is in the state \CLOSED{} and the other (TCP B) is in the state \LISTEN.
    \begin{enumerate}
        \item TCP A wants to initiate a connection with TCP B. It sends a SYN segment to
            TCP B and passes to the state \SYNSENT{}.

        \item TCP B receives a segment SYN. It has to respond to this segment by a segment
            containing the flags SYN and ACK (acknowledgment of the received segment).
            It changes its state for \SYNRECEIVED.

        \item When TCP A receives the segment sent by TCP B containing the flags SYN and ACK,
            it only has to send back an ACK of this segment to pass in the state \ESTABLISHED.

        \item TCP B receives the ACK of its segment and pass in state \ESTABLISHED{} too.
    \end{enumerate}
    At the end of this procedure, the two TCPs are in state \ESTABLISHED{} and can start to
    send data. Figure \ref{fig:handshake} illustrates the procedure described above. This procedure to open a connection is known as the ``three-way
    handshake'' in the TCP norm.

    \begin{figure}
        \centering
        \begin{tikzpicture}
            \node (A) at (0, 5) {\bfseries TCP A};
            \node (B) at (5, 5) {\bfseries TCP B};
            \draw[dashed] (-.7,4.8) -- (5.7,4.8);
            \node (TCPA1) at (0, 4.4) {\CLOSED{}};
            \node (TCPB1) at (5, 3.9) {\LISTEN{}};
            \node (TCPA2) at (0, 3.4) {\SYNSENT};
            \node (TCPB2) at (5, 2.9) {\SYNRECEIVED{}};
            \node (TCPA3) at (0, 2.4) {\ESTABLISHED{}};
            \node (TCPB3) at (5, 1.9) {\ESTABLISHED{}};

            \draw[->] (TCPA1) -- (TCPA2);
            \draw[->] (TCPB1) -- (TCPB2);
            \draw[->] (TCPA2) -- (TCPA3);
            \draw[->] (TCPB2) -- (TCPB3);
            \draw[->, dashed] (0.2, 3.9) -- node [midway, fill=white] {\footnotesize SYN} (TCPB1);
            \draw[->, dashed] (4.8, 3.4) -- node [midway, fill=white] {\footnotesize SYN + ACK} (TCPA2);
            \draw[->, dashed] (0.2, 2.9) -- node [midway, fill=white] {\footnotesize ACK} (TCPB2);
        \end{tikzpicture}
        \par
        \caption{Three-way hanshake procedure. The solid lines represent the transition between
        the states and the dashed lines represent the messages sent.}
        \label{fig:handshake}
    \end{figure}

    \bigskip

    We want to point out to the reader the differences that exist between the state machine
    of figure \ref{fig:TCPAutomaton} and the one that can be found in~\cite[23]{rfc793}. In particular
    a transition between the states \FINWAITONE{} and \TIMEWAIT{} has been added. It reflects the case where
    the ACK and FIN flags are received in the same segment. It is an allowed behavior by the norm, and
    also it is closer to what is done in CycloneTCP library.

    For more details on how two TCPs can establish or close a connection, the reader can refer
    to the TCP norm where many scenarios are described and explained, and more
    details are given on the state and segments that can be sent.

    \subsubsection{Multitasking in the implementation}

    Different tasks can interact with a socket. The TCP norm gives an example with
    three tasks: one for the \emph{user calls}, one for the \emph{arriving segments}
    and one for the \emph{timers}. This design has also been adopted in CycloneTCP.

    \paragraph{User calls}
    User calls refer to functions that can be called by the user to control the connection,
    send, or receive data: OPEN, CLOSE, ABORT, SEND, RECEIVED. Transition between states of
    the connection can happen during the call of these functions since they are intended to control
    the connection.

    \paragraph{Arriving segments}
    In this task, the received segments are processed. The corresponding messages are sent back.
    Transitions between states can happen here when a message is received. For example,
    this is the case when a message containing a SYN is received when the socket is in
    the \LISTEN{} state. It will automatically send a SYN + ACK in response, and change
    its state for \SYNRECEIVED.

    \paragraph{Timers}
    The timers control the timeouts like the retransmission timeout to retransmit a message, or
    the time-wait timeout to close the connection after an amount of time. Then, transitions can also
    happen in this task.

    To sum up, a single socket can be seen as a shared structure over multiple tasks, that can all
    manipulate it and change the value of its fields. The schema in figure \ref{fig:TCPConcurrency} illustrates that.

    Only one task can perform an operation on the socket at the same time. The access to the fields of
    the socket are protected by mutex.

    Two tasks can communicate synchronously or asynchronously together. The synchronous communications are based on
    the interface provided by the OS, in particular with events.

@@

            \draw[decorate, decoration={brace, amplitude=10pt}] (7.5, 0) -- (7.5, -1.8)
                node [black,midway,xshift=0.55cm, rotate=-90] {\textit{Tasks}};

@@

    The TCP norm defines a lot of behavior that could be verified by automatic proof.
    Here is a non exhaustive list of what could be done:
    \begin{itemize}
        \item Verification of the transitions between the states.
        \item Integrity of the messages sent, \textit{i.e.} we could check that the sent
        message contains the correct flag and the correct sequence and acknowledgment numbers.
        \item Integrity of the messages received, \textit{i.e.} we could check that the
        messages received are correctly processed in regard of the flags they contain.
        \item Functional correctness of the user functions in regard to their specifications.
    \end{itemize}

    Some choices have had to be made in regard to the short time devolved to the internship, and
    not every aspect of the TCP norm has been proved yet. However, some aspects have been selected
    for the challenge or the importance they represent. A list of these aspects is presented
    here and the choices will be motivated in the next subsections. First of all, we have wanted
    to improve the interface of the functions. The other point was to improve the safety of the
    protocol itself.

@@

    One of the main problems pointed to by the main author of the CycloneTCP library is that users
    do not correctly use the API supplied. As a result, some function
    calls can be incorrect, either because the arguments are incorrect, or because the
    return code of the previous calls has not been checked, which leads to an
    incorrect code.

@@

    The multiple tasks and the changes that can happen at every moment in one
    task or another make the verification hard. Multiple interactions exist
    between the multiple tasks: synchronous and asynchronous.
    All these interactions have to be considered if we want to write correct contracts
    in regard to what could be done in other tasks.

    As SPARK does not have a native mode to deal precisely with interactions related to concurrency, we have to
    model these different interactions by hands by writing assertions that
    modify the state of the socket as another task could do it. In particular
    the problem has been encountered when we tried to write contracts for the correctness
    of the user functions. More details and explanations will be given in the following
    section, where we will investigate the solution found to these technical problems.

@@

    This part mainly focuses on the high level user functions, the socket functions.
    The socket functions are the one called by the user to perform operations on
    the network. They are located in the files \texttt{socket\_interface.ad(b|s)}.

@@

    In order to verify the transitions we have read the norm to extract information about
    the allowed transitions, and we have added contracts in the function
    \lstinline[language=Ada]{tcpChangeState}. Since all the transitions are made through
    this function, an incorrect transition will lead to a message by SPARK.

    As mentioned, since most transitions are done outside the user functions, a big part
    of this work was to consider other transitions in other tasks and we present how we have
    done it in the following sections.

@@

    Almost all the transitions are done in the file \texttt{tcp\_fsm.c}. This file has not been
    translated in SPARK by a lack of time. Then we do not have strong guarantees on its functions
    behavior. However, for formal verification we need to know what is done in the functions of
    this file, because a wrong contract can make all the verification wrong.

@@

    All the work related to KLEE is present in the folder \texttt{klee/} and can be compiled and run
    thanks to the makefile to reproduce the results. Roughly, what is done is creating a random
    incoming message, put the socket in the desired state and call the desired function to see what
    final states can be obtained.

    Finally, all the results found by KLEE have been reported as a post-condition of the function \lstinline[language=Ada]{Tcp_Process_One_Segment}.
    This function is essential for all the verification of concurrent parts and is reused everywhere
    the concurrency is involved to compute possible interferences. It follows that the safety of the code
    relies on the confidence we have in this function.

@@

    Let $\rightarrow$ be the transition function in the TCP automaton restricted to the transitions
    that can be performed by the reception of a message and its automatic response mechanism.
    Then we have $\rightarrow = \text{\lstinline[language=Ada]{Tcp_Process_One_Segment}}$ and
    we need to consider the reflexive transitive closure $\rightarrow^*$ of $\rightarrow$, with
    \[\rightarrow^* = \bigcup_{n\in\mathbb{N}} \rightarrow^n\]
    Now, by examining the TCP automaton in figure \ref{fig:TCPAutomaton}, we see that all state is
    only reachable by a state at a distance less than 3 (the maximum path is between \SYNSENT{} and
    \CLOSEWAIT{} if we pass by the states \SYNRECEIVED{} and \ESTABLISHED{}) without user action. Since
    we only consider the transitive closure in term of states, we can significantly reduce the number
    of iteration of $\rightarrow$ to compute $\rightarrow^*$ and finally we get:
    \[\rightarrow^* = \bigcup_{n=1}^3 \rightarrow^n\]
    It follows that algorithm \ref{algo:tcpProcessSegment} is good enough to compute the transitive closure of the function
    \lstinline[language=Ada]{Tcp_Process_One_Segment} by taking advantage of the fact that SPARK can unroll
    small loops to obtain more precise results. The contract of \lstinline[language=Ada]{Tcp_Process_Segment} is
    manually written and proved with SPARK.

@@

    In the code, the function \lstinline[language=C]{tcpUpdateEvents} is called at different locations; it
    updates the events true for the socket and it can raise the desired event.
    In particular this function is called each time the state of
    the socket is changed. This is sufficient for our purpose since we are only interested in the state
    changes.

@@

    \begin{algorithm}[t]
        \SetKwFunction{ProcessSegm}{\lstinline[language=Ada]{Tcp\_Wait\_For\_Events}}
        \SetKwProg{Fn}{function}{}{end}
        \Fn{\ProcessSegm{Socket, Event, Event\_Mask}}{
            $S_{last} := \text{\textit{Socket}}$\;
            $S := S_{last}$\;
            $E :=$ \lstinline[language=Ada]{Tcp_Update_Events}($S_{last}$)\;
            \If{$(E\ \&\ \text{Event\_Mask}) \neq 0$}{
                \KwRet{$S$}\;
            }
            \For{$i=1$ \KwTo $3$}{
                $S_{last} :=$ \lstinline[language=Ada]{Tcp_Process_One_Segment}($S_{last}$) \;
                $S := S \cup S_{last}$\;
                $E :=$ \lstinline[language=Ada]{Tcp_Update_Events}($S_{last}$)\;
                \If{$(E\ \&\ \text{Event\_Mask}) \neq 0$}{
                    \KwRet{$S$}\;
                }
            }
            \KwRet{$\emptyset$}\;
        }
        \caption{Function to compute the possible state after when waiting for a particular event.}
        \label{algo:waitForEvents}
    \end{algorithm}

@@

    \textcolor{red}{\paragraph{Weakness} If two tasks want to lock the same mutex at the same time, which one wins?
    This is probably something to discuss with Clément, because the function \texttt{tcpWaitForEvents} could have a
    different semantic than what we imagine at first.}

    \subsection{Bug found}

    Thanks to this work, a bug has been found. The contract puts on the function \lstinline[language=Ada]{Tcp_Change_State}
    warned of an unauthorized transition. We reproduce the code that leads to this transition and we will explain
    why it is a bug.
    \begin{lstlisting}[language=Ada, basicstyle=\ttfamily\small, numbers=left, numberstyle=\tiny, escapechar=\%]
   case Sock.State is
      when TCP_STATE_SYN_RECEIVED
         | TCP_STATE_ESTABLISHED =>
      -- Flush the send buffer
      Tcp_Send (Sock, Buf, Ignore_Written,
                SOCKET_FLAG_NO_DELAY, Error);
      if Error /= NO_ERROR then
         return;
      end if;

      -- Make sure all the data has been sent out
      Tcp_Wait_For_Events % \label{bugProg:tcpWaitForEvents} %
         (Sock       => Sock,
          Event_Mask => SOCKET_EVENT_TX_DONE,
          Timeout    => Sock.S_Timeout,
          Event      => Event);

      -- Timeout error?
      if Event /= SOCKET_EVENT_TX_DONE then
         Error := ERROR_TIMEOUT;
         return;
      end if;
      % \label{bugProg:tcpWaitForEventsPost} %
      -- Send a FIN segment
      Tcp_Send_Segment
         (Sock         => Sock,
          Flags        => TCP_FLAG_FIN or TCP_FLAG_ACK,
          Seq_Num      => Sock.sndNxt,
          Ack_Num      => Sock.rcvNxt,
          Length       => 0,
          Add_To_Queue => True,
          Error        => Error);
      -- Failed to send FIN segment?
      if Error /= NO_ERROR then
         return;
      end if;

      -- Switch to the FIN-WAIT-1 state
      Tcp_Change_State (Sock, TCP_STATE_FIN_WAIT_1);
    \end{lstlisting}

    The function \lstinline[language=Ada]{Tcp_Send} will change the state of the socket for
    the state \ESTABLISHED{} or \CLOSEWAIT{} in our case, if no error happens. Moreover the next
    function at line \ref{bugProg:tcpWaitForEvents} has a contract that contains:
    \begin{lstlisting}[language=Ada, basicstyle=\ttfamily\small]
(if (Event_Mask and SOCKET_EVENT_TX_DONE) /= 0 then
   (if Event = SOCKET_EVENT_TX_DONE then
      -- RST segment received
      Model(Sock) = (Model(Sock)'Old with delta
         S_State => TCP_STATE_CLOSED,
         S_Reset_Flag => True) or else
      (if Sock.S_State'Old = TCP_STATE_CLOSE_WAIT then
         Model(Sock) = Model(Sock)'Old or else
      elsif Sock.S_State'Old = TCP_STATE_ESTABLISHED then
         Model(Sock) = Model(Sock)'Old or else
         Model(Sock) = (Model(Sock)'Old with delta
            S_State => TCP_STATE_CLOSE_WAIT))))
    \end{lstlisting}
    which means that at line \ref{bugProg:tcpWaitForEventsPost}, the socket can either be
    in state \ESTABLISHED{}, \CLOSEWAIT{} or \CLOSED{} if the connection had been reset by the remote
    during the call of \lstinline[language=Ada]{Tcp_Wait_For_Events}, when the mutex was released.
    The precondition of the function \lstinline[language=Ada]{Tcp_Change_State} encodes the allowed
    transitions that we can see in the automaton in \ref{fig:TCPAutomaton}, and thus the transition
    \CLOSEWAIT{} $\rightarrow$ \FINWAITONE{} is not allowed, nor the transition \CLOSED{} $\rightarrow$ \FINWAITONE{}.

    This incorrect behavior has been detected thanks to the use of Klee and the improvement
    on how function's closure are computed. The contract for the call
    \begin{lstlisting}[language=Ada, basicstyle=\ttfamily\small]
    Tcp_Wait_For_Events (Sock, SOCKET_EVENT_TX_READY,
                         Sock.S_Timeout, Event);
    \end{lstlisting}
    was not correct at first, because it was written by hands and thus not enough precise.
    Then, the possible state at the end of the function \lstinline[language=Ada]{Tcp_Change_State}
    was reduce to \ESTABLISHED{} and the same problem existed for the call
    at line \ref{bugProg:tcpWaitForEvents}. At this point no problem was detected with SPARK,
    because the contracts were true. With Klee, some transitions that were not considered at first,
    because hide in subfunctions have been discovered. In particular there exists a transition from
    \SYNRECEIVED{} to \CLOSEWAIT{} if a FIN flags is present in the same segment of the ACK that was not
    spotted before. Putting these information in the function \lstinline[language=Ada]{Tcp_Process_One_Segment},
    and proving the post-condition of the function \lstinline[language=Ada]{Tcp_Wait_For_Events}
    directly with SPARK using algorithm \ref{algo:waitForEvents}, rather than not proving it like
    it was done before, helps to correct the contracts and find the bug.

@@

    Other improvements that could be done in the future (this is a not exhaustive aggregate of ideas,
    this list is subject to change):
    \begin{itemize}
        \item Fully translate the file \texttt{tcp\_misc.c} in Ada to have more guarantees.
        \item Same for the file \texttt{tcp\_fsm.c} that has been investigated with klee. But
        it is not enough and we need more guarantees for this file, in order to prove the
        interactions between tasks first, and also to prove the correctness of this part
        of code w.r.t. the TCP norm.
        \item The translation only considers one possible preprocessing. A future work would
        be considering all the possible preprocessing by adding constants like in the C
        version. An harder work is to find a way to keep a coherence between the C preprocessing
        and the Ada preprocessing. It can be achieved with an automatic tool, for example.

        \item It is not checked that the correct flags have been sent before changing of state.
        It could be added by using ghost variables, or abstract states.
    \end{itemize}
